A User-Centric Approach to the Design and Consequences of Recommender Systems

### 2019-10-01 paper/filter_bubbles.tex
remove unused package
updated paper

\documentclass[sigconf]{acmart}
\settopmatter{printacmref=false}
\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage[font=small]{caption}
\usepackage{float}
\newtheorem{finding}{Finding}

\usepackage{xspace}
\newcommand{\OMIT}[1]{}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% Rights management information. 
% This information is sent to you when you complete the rights form.
% These commands have SAMPLE values in them; it is your responsibility as an author to replace
% the commands and values with those provided to you when you complete the rights form.
%
% These commands are for a PROCEEDINGS abstract or paper.
\copyrightyear{2019}
\acmYear{2019}
\setcopyright{acmlicensed}
\acmConference[WWW '20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20--24, 2020}{Taipei, Taiwan}
\acmBooktitle{WWW '20: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmDOI{10.1145/1122445.1122456}
\acmISBN{978-1-4503-9999-9/18/06}

@@

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{The Ex-Ante View of Recommender System Design}
%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
\end{abstract}

@@

%
% Keywords. The author(s) should pick words that accurately describe the work being
% presented. Separate the keywords with commas.
\keywords{Recommender Systems, Beliefs, Decision Theory, Filter Bubbles}

@@

Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets. For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube. users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption. Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.

Recommender systems have been influential in shaping user choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations. While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead users into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.

\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of user decision-making in the context of large choice sets faced by users in online markets. We then ask how a stylized model of recommendation can shape the behavior of long-lived users and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.

We focus on understanding the degree to which recommender systems can lead users into \textit{filter bubbles}. Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}. The key component of our model, drawn from the results of \cite{schulz2019structured}, is that the utilities of similar goods are correlated. This means that when users consume a good and learn its true utility this gives them information about similar goods. Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty. This learning spill-over induces users to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns. This effect is further exemplified when users are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.

We then consider a stylized model of user recommendation and explore how recommendation can affect consumption patterns. We suppose each good is a sum of a common value component and an idiosyncratic component, where the idiosyncratic component is inherently unpredictable given other individual's preferences. We model existing recommendation systems as providing users with information on the common-value component and denote as this as partial recommendation. We consider how user behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for users if they knew the true utility of all the goods in their choice set.

Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity. However, we also find that partial recommendation increases homogeneity amongst users.

\textbf{Related Work.}
Recommender Systems

@@

\subsection{Preliminary on Expected Utility Theory}
[Duarte can do this]

\subsection{Model Setup}
\par
\noindent \textbf{users} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal{N} = \left\{0,1,...,N-1\right\}$. For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{N}$ the first time they consume it.

We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$. In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner:
\begin{align*}
u_{in}= v_{in} + \beta v_n
\end{align*}
where $v_{in}$ denotes an idiosyncratic component -- i.e. user $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{+}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta \rightarrow \infty$, every individual has the same preferences. [include some discussion of what this implies about the predictability of a user's preferences]

Stacking utilities in vector-form, we get:
\begin{align*}
{\left(u_{in}\right)}_{n \in \mathcal{N}}=:U_i=V_i+ \beta V
\end{align*}
We stack the idiosyncratic and common-value components as $V_i ={\left(v_{in}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
\par
\noindent \textbf{user Beliefs} We assume that users do not know the realized utilities of the goods before consuming an item.  Formally, user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal. The idiosyncratic utility and common-value utility are distributed as follows:
\begin{enumerate}
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
\end{enumerate}

We impose the normality assumption for two reasons. The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates. The second is that it allows us to incorporate an easily interpretable correlation structure between the items.

Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in \mathcal{N}}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d. These $\overline v_{in}$ denote the prior belief that $i$ holds about the her valuation over good $n$. As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{in}$ denotes this idiosyncrasy at the level of prior beliefs.
\par
\noindent \textbf{user Learning}
When a user consumes a good $n$ they learn the realized utility for that good. In our model we incorporate the idea that learning the utility of good $n$ gives a user information about similar items. We assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.

In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as:
\begin{align*}
d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}
\end{align*}
 $m$ and $n$ are the indices of the items in $\mathcal{N}$. We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$. The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby. Informativeness, for any $\rho \in (0,1)$, is decreasing in distance. The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\rho$ , $(n,n+2)$-th entry is $\rho^2$, etc.
\par
\textbf{user Decision-Making}
We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$. This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it. For tractability we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs. Any ties are broken uniformly at random. This assumption is critical not only for tractability but also in order to have easily defined benchmarks.
\par
\noindent \textbf{Recommendation}
Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of looking at realistic implementations of recommender systems. We model recommendation as giving users information about the utility of the goods.

We will consider three recommendation regimes:
\begin{enumerate}
\item Partial Recommendation
\item No Recommendation
\item Omniscient Recommendation
\end{enumerate}
The case of primary interest is \textit{partial recommendation} where the recommender observes utilities accrued, but does not know user $i$'s starting beliefs $\overline U_i$. In this case, the recommender knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ before recommendation.} However, the recommender does know the users' beliefs, $\overline V_i$. Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\overline V_i$. User's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow recommendation.

The notion of partial recommendation that we consider is idealized where the recommendation does the Bayesian updating for users. An alternative that is equivalent would be the recommender simply providing $V$ without knowing $\overline V_i$. In this case, the recommendation will be that the user $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{N} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$. For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
The best item that could recommended with such information if the item with highest common value component. However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component. Therefore, we assume that the RS reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
} The results would be equivalent, but this would shift the burden of Bayesian updating to users.

We further consider two cases that serve primarily as benchmarks. The first is \textit{no recommendation}, where users get no additional information about utilities and make consumption choices based on their beliefs and consumption history. This gives us a benchmark as to how users in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation. The second is \textit{omniscient recommendation} where the recommender knows the true realized utility of each good for each user and can therefore recommend the best remaining good in every period. This gives us a full information benchmark, which is the optimal consumption path for a user if all uncertainty about their preferences was resolved.
\par
\noindent \textbf{Simulation Details}. We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically. The Gaussian assumption allows for closed form Bayesian-updating which enables us to simulate our model but does not provide much help in analytical characterizations.

We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations. We simulate our model over $100$ populations of users with $100$ users per population. For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population. A given set of parameters and a population are a single data point in our dataset.

We simulate over a grid of $\gamma, \sigma, \rho, \beta$. When we consider results varying a single parameter, we group the results over the other parameters and consider the differences in value only varying the parameter of interest. We report results for a relatively small $T$. Unless otherwise reported, our results are robust to varying $N$.

\subsection{Discussion of Assumptions}

In this section we provide further discussion of several of the assumptions of our model.

\noindent \textbf{Uncertainty}: We assume that users have underlying uncertainty about the realized utility of the items in their choice sets and as a result face a decision problem under uncertainty. The first justification for this is that environments where recommender systems are commonly deployed, such as on media streaming or e-commerce platforms, contain experience goods. A common assumption in the economics literature is that for experience goods, where users can only observe their realized utility after consumption, users face a decision problem under uncertainty when making consumption choices. The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets.\footnote{Recent empirical work [cite the cameras paper and hodgson-lewis] has shown that, for product search, users engage in a ``zooming in" search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space.} Even if they acquired all the information from product reviews, this would only fully resolve uncertainty if individuals did not have idiosyncratic tastes.

This motivates modeling recommendation as providing users \textit{information} in order to reduce their uncertainty.

\noindent \textbf{Large Choice Sets}: We are interested in environments where the number of items in the choice set is very large and users only consume a small number of these items. This is realistic for platforms such as Netflix, Spotify, Amazon, and YouTube where there are sometimes billions of items and users can realistically only consume a handful of these items in their lifetime.

\noindent \textbf{Learning Spillovers}: Users in our environment face a sequential decision-making problem. However, we suppose that individuals are myopic (i.e. consume the item with the highest expected utility, given their beliefs in a single period) in order to maintain tractability. The predominant difficulty in the learning problem they face is not that they need to repeatedly sample a single item in order to learn its value (as in standard multi-armed bandit formulations), but rather that they need to be able to generalize from their experience from consuming one item to other items. After consuming a single item they learn the realized utility of the good.

\indent \cite{schulz2019structured} studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders. They find that individuals engage in similarity-based generalization where learning about the realized utility of a particular good provides them with information about similar goods. Content-based recommender systems exploit a similar idea where, if a user liked one item, then it is also likely that they will like a similar, but different, item as well.

Our model incorporates this observation in a stylized fashion where the degree of information spill-overs is controlled via parameter $\rho$ and we study how the strength of the spill-overs impacts behavior and the effect of recommendation.

\section{Results}

\subsection{Local Consumption}

>>>>

\documentclass[sigconf, anonymous, review]{acmart}
\settopmatter{printacmref=false}
\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage{float}
\usepackage{enumitem}
\newtheorem{finding}{Finding}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% Rights management information. 
% This information is sent to you when you complete the rights form.
% These commands have SAMPLE values in them; it is your responsibility as an author to replace
% the commands and values with those provided to you when you complete the rights form.
%
% These commands are for a PROCEEDINGS abstract or paper.
\copyrightyear{2020}
\acmYear{2020}
\setcopyright{acmlicensed}
\acmConference[WWW'20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20-24, 2020}{Taipai, Taiwan}
\acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmDOI{10.1145/1122445.1122456}
\acmISBN{978-1-4503-9999-9/18/06}

@@

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{A User-Centric Approach to the Design and Consequences of Recommender Systems}
%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
We study a model of user decision-making in the context of recommender systems. We explore the extent to which issues traditionally associated with recommender systems -- namely filter bubbles, user homogenization and consumption diversity -- can be attributed to recommendation systems. The key component of our model is that users face a decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty about similar products. We find that this generates behavior consistent with narrowing filter bubble effects and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al. (2014). However, we find that recommendation, while alleviating filter bubble effects, leads to an increase in user homogeneity.
\par

We discuss the evaluation of recommender systems in light of these findings. Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved. We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, that is, information without which choice would be different and lead to worse outcomes on average relative to the case where no recommendation is provided. As a result, we emphasize the importance of understanding user behavior without recommendation, both for designing recommendations and understanding their impacts and usefulness.
\end{abstract}

@@

%
% Keywords. The author(s) should pick words that accurately describe the work being
% presented. Separate the keywords with commas.
%\keywords{Recommender Systems, Filter Bubbles, Serendipity}

@@

Recommender Systems (RS) have become critical for assisting consumers in navigating the large choice sets that they face in many online markets. For instance, consumers have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube. Consumers in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption. Furthermore, consumers interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
\par

Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{\url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/} that personalized recommender systems lead consumers into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead consumers to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
\par

\textbf{Our Model and Contributions} In this paper, 
%drawing from recent work in psychology and decision theory in economics, 
we first propose a model of consumer decision-making in the context of large choice sets faced by consumers in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations. 
We then ask how a stylized model of recommendation can shape the behavior of long-lived consumers 
and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
\par

The key components of our model are:
\begin{enumerate}[label=(\arabic*)]
\item Users face uncertainty about their valuation of the different goods and thus face a decision problem under under uncertainty.
\item Choice sets are large and users only consume a small fraction of this choice set over their lifetime.
\item Consumption of these items is sequential and users live for multiple periods.
\item Consuming an item provides information about the value of similar items in the product space.
\item Recommendation provides users with information, reducing their uncertainty.
\end{enumerate}

Using this model we first investigate the degree to which recommender systems lead users into filter bubbles. Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that in the context of movie consumption, user behavior is consistent with filter bubble effects \textit{even without recommendation}. The driving force of this in our model is that utilities of similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods. Crucially, this not only impacts the underlying belief about the average utility of the good, but also the level of uncertainty. Consequently, this learning spillover induces users to consume goods ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space. This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
We compare this benchmark to the case where recommendations are provided. We consider recommendations as giving information to consumers in order to reduce their uncertainty and we find that this reduces the narrowing effect, which is also consistent with the empirical results from \cite{nguyen2014exploring}.
\par

We then explore how diverse the overall set of items that users consume is and find that, without recommendation, higher diversity of items does not ensure higher welfare.
This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of utilities, leading to higher overall value from consumption. When they consume products they value little, they move to another part of the product space. A sequence of bad realizations will then lead to very intense search over very different parts of the product space and high consumption diversity, whereas discovering a sweet spot leads to highly rewarding local search and very low diversity. Thus we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users. 
\par

We find that recommendation can lead to increases in homogeneity among users. We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component. This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommonder can learn from previous users' data. The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation. This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component, resulting in an increase in homogeneity.
\par

Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems. Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without the recommendation. Thus, in order to evaluate the value of recommendation, it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior. If recommendations fail to affect behavior, they are of no use: the user would consume the same good and learn its value with and without the recommendation. Moreover, we argue that successful recommendation systems have to take into account the degree of uncertainty users have over different portions of the product space. Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in face of risk-aversion.
\par

\textbf{Related Work.} 
\cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content. A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}. The most relevant to our proposed study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption. They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume. To our knowledge there are no theoretical models that rationalize the empirical findings but predominantly empirical studies. Our paper provides a theoretical framework through which to view this problem.
The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other. \cite{celam2008hits,treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization. \cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems. Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume goods in those regions.
\cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here. \cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of utilities in the environment, but consider it in the context of search for a single item.
Our discussion of recommender system evaluation contributes with insights closely tied to the value of information that is used in the economics literature (see e.g. \cite{bergemann2019information}) and builds off the large literature stemming from \cite{mcnee2006being} that makes the case that accurate predictions of true realized valuations may not lead to the best recommendations. This is related to the notion of serendipity in recommender systems that has been studied extensively (e.g. see \cite{kotkov2016survey} for a survey).

@@

\subsection{Preliminaries on Expected Utility Theory}
\noindent For every product $n$ in the product space $\mathcal J=\{0,1,....,N-1\}$, we assume that each consumer $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it. Each consumer can value the same product in differently. However, we assume that consumers have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous. So, ex-post, the value of product $n$ for consumer $i$ is given by $u(x_{i,n})$ However, before consuming the product, the consumer does not know exactly how she will value it. In particular, even consumers that will end up having the same ex-post valuation of product $n$ may differ in their ex-ante valuation because they hold different beliefs about it. We denote by $p_{i}$ the beliefs consumer $i$ has about how she will value each of the products in the product space. Note that this implies that consuming product $n$ is the same as taking a gamble. Each individual evaluates the product according to the associated expected utility associated with the product, i.e. $U_i(n)=\mathbb E_{p_i}[u(x_n)]$. 
\par

Risk aversion captures the how different individuals react to the risk associated to a particular consumption opportunity. It is formalized as follows: Suppose for a given gamble $x$ taking real values and following distribution $p$, there is a certain amount of money that leaves the individual indifferent between taking the gamble and taking this sure amount, call it $\delta(x)$, termed the \textit{certainty equivalent} of $x$. An individual $i$ is more risk-averse than another individual $j$ when for every gamble $x$, individual $i$ requires a lower value to give up the gamble when compared to individual $j$, i.e. $\delta_i(x)$. Therefore, a more risk-averse individual is more willing to avoid the risk of taking the gamble. We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences. Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
\par

\subsection{Model}
\par
\noindent \textbf{Users} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$. For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
\par

We denote by $x_{i,n}$ individual $i$'s realized value from consuming item $n$. In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e. user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much product $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
\par
Stacking utilities in vector-form, we get the vector of values associated with each item $${\left(x_{i,n}\right)}_{n \in \mathcal{N}}=:X_i =V_i+ \beta V $$ where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
\par
\noindent\textbf{User Decision-Making}
We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$. This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it. Since this is a sequential decision-making problem under uncertainty, in principle consumers face an exploration-exploitation trade-off. However, for tractability, we impose that consumers are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
\par
\noindent \textbf{User Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item.
Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal
\begin{enumerate}
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
\end{enumerate}

We impose the normality assumption for two reasons. The first is that consumers update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates. The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
\par
Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed. These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about the her valuation over good $n$. As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
\par

We assume users are expected utility maximizers. User $i$'s certainty equivalent for product $n$, the sure value that makes user $i$ indifferent between it and consuming the product $n$, conditional on the consumption history, is given by
$\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$.

\noindent \textbf{User Learning}
When a consumer consumes a product $n$ they learn the realized value for that product. In our model, we incorporate the idea that learning the value of product $n$ gives a consumer information about similar items. 
%This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
We assume that learning about the value of good $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
\par

In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$. This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior. We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$. The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby. Informativeness, for any $\rho \in (0,1)$, is decreasing in distance. The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\rho$, the $(n,n+2)$-th entry is $\rho^2$, etc.
\par
\begin{center}
\includegraphics[width=.4\linewidth]{Example-Bubbles.pdf}
\captionof{figure}{Illustrative Example}
\label{fig:illustrative_example}
\end{center}

\noindent \textbf{An Illustrative Example} We illustrate our model with a simple example. Suppose that there are four products: 0, 1, 2, 3. The products are in different places of the product space, where 1 is close to 0 and 2 but more distant from 3, as shown in Figure \ref{fig:illustrative_example}.
Suppose that the user actually believes that they all have expected value of 0 and consumes product 1. The underlying correlation structure implies that upon observing that $x_1=y$ the user will update beliefs about $\mathbb E x_0$, $\mathbb E x_2$ and $\mathbb E x_3$ to $\rho y$, $\rho y$ and $\rho^2 y$ respectively. This is because learning about product 1's value is more informative about similar products' value, 0 and 2, than products further away in the product space, 3. Therefore, the user will adjust her beliefs about their mean -- zero in the example -- more strongly for products which are similar. Moreover, now uncertainty about the value of products 0, 2 and 3 is also decreased but differentially so: learning product 1's value has contributed to a stronger reduction of uncertainty for nearby products that for product 3. If the actual value was positive, this induces the user to perform some local exploration, as 0 and 2 have higher mean and lower variance. If instead it was negative, then user's posterior beliefs penalize more strongly products 0 and 2 than 3, so if she is not too risk averse, she would go across the product space in search of some better element. However, if risk aversion is high enough, the user might be willing to trade-off ex-ante expected lower values for lower risk and stick to local exploration of products that have just because the user has less uncertainty about their values. As the saying goes, if you are risk-averse, better the devil you know.
\par

\noindent \textbf{Recommendation}
Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice. We model recommendation as giving consumers information about the valuation of the products.
\par

We will consider three cases. The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know neither $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision.} However, the recommender does know the users' beliefs $\bar V_i$. Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$. Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.
\par
